Packaging research artefacts with RO-Crate

### 2022-01-02 latex/ro-crate.tex
URI->IRI.  license() is also a relation.

The possibilities of consuming\footnote{Some consideration is needed in processing of RO-Crates as
knowledge graphs, e.g. establishing absolute URIs for files inside a
ZIP archive, detailed in the RO-Crate specification: \url
{https://www.researchobject.org/ro-crate/1.1/appendix/relative-uris.html}.} RO-Crate metadata with such
powerful tools gives another strong reason for using Linked Data as a
foundation. This use of mature Web\footnote{Note that an RO-Crate is not required to be published on the
Web, see Section~\ref{sec:selfdescribed}.} technologies also means its
developers and consumers are not restricted to the Research Object
aspects that have already been specified by the RO-Crate community, but
can extend and integrate RO-Crate in multiple standardised ways.

@@

One aim of RO-Crate is to be conceptually simple. This simplicity has
been repeatedly checked and confirmed through an informal community
review process. For instance, in the discussion on supporting ad-hoc
vocabularies\footnote{\url{https://github.com/ResearchObject/ro-crate/issues/71}} in
RO-Crate, the community explored potential Linked Data solutions. The
conventional wisdom in RDF best
practices\footnote{\url{https://www.w3.org/TR/swbp-vocab-pub/}} is to establish a
vocabulary with a new URI namespace, formalised using RDF
Schema\footnote{\url{http://www.w3.org/TR/2014/REC-rdf-schema-20140225/}} or
OWL\footnote{\url{http://www.w3.org/TR/2012/REC-owl2-overview-20121211/}}
ontologies. However, this may seem an excessive learning curve for
non-experts in semantic knowledge representation, and the RO-Crate
community instead agreed on a dual lightweight approach: (i)
Document\footnote{\url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html\#adding-new-or-ad-hoc-vocabulary-terms}}
how projects with their own Web-presence can make a pure HTML-based
vocabulary, and (ii) provide a community-wide PID namespace under
\texttt{https://w3id.org/ro/terms} that redirect to simple CSV files
maintained in GitHub.\footnote{\url{https://github.com/ResearchObject/ro-terms}}

@@

RO-Crate reuses vocabulary terms and IRIs from Schema.org, but provides
its own versioned JSON-LD
context,\footnote{\url{https://w3id.org/ro/crate/1.1/context}} which has a flat list
with the mapping from JSON-LD keys to their URI equivalents (e.g. key
\texttt{"author"} maps to the \url{http://schema.org/author} property).

@@

An $\textit{ROCrate}(R)$ is defined as a self-described \emph{Root Data Entity},
which describes and contains parts (\emph{data entities}), which are
further described in \emph{contextual entities}. These terms align with
their use in the RO-Crate 1.1 terminology.\footnote{\url{https://www.researchobject.org/ro-crate/1.1/terminology}}

The $\textit{Root}(r)$ is a type of $\textit{Dataset}(r)$, and must have the metadata to
literal attributes to provide a $\textit{name}$, $description$ and
$\textit{datePublished}$, as well as a contextual entity identifying its
license. These predicates correspond to the RO-Crate 1.1 requirements
for the root data entity.\footnote{\url
{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html\#direct-properties-of-the-root-data-entity}}

@@

The below is an example RO-Crate represented using the above
formalisation, assuming a base URI of \texttt{<http://example.com/ro/123/>}:
%
{\fontsize{9.7}{11.7}\selectfont{\begin{gather*}
\textit{ROCrate}(\texttt{<http://example.com/ro/123/>})
\\
\textit{name}\bigl(\texttt{<http://example.com/ro/123/>},
\\
\quad \text{``Data files associated with the manuscript:Effects of {
\ldots}''}\bigr)
\\
\textit{description}\bigl(\texttt{<http://example.com/ro/123/>},
\\
\quad \text{``Palliative care planning for nursing home residents {
\ldots}''}\bigr)
\\
\textit{license}(\texttt{<http://example.com/ro/123/>},
\\
\quad \texttt{<https://spdx.org/licenses/CC-BY-4.0>})
\\
\textit{datePublished}\bigl(\texttt{<http://example.com/ro/123/>}, \text
{``2017-02-23''}\bigr)
\\
\textit{hasPart}(\texttt{<http://example.com/ro/123/>}, \texttt{<http://example.com/ro/123/file.txt>})
\\
\textit{hasPart}(\texttt{<http://example.com/ro/123/>}, \texttt{<http://example.com/ro/123/interviews/>})
\\
\\
\textit{ContextualEntity}(\texttt{<https://spdx.org/licenses/CC-BY-4.0>})
\\
\textit{name}\bigl(\texttt{<https://spdx.org/licenses/CC-BY-4.0>},
\\
\quad \text{Creative Commons Attribution 4.0''}\bigr)
\\
\\
\textit{ContextualEntity}(\texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>})
\\
\textit{name}\bigl(\texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>},
\\
\quad \text{Creative Commons Attribution Non Commercial 4.0''}
\bigr)
\\
\\
\textit{File}(\texttt{<http://example.com/ro/123/survey.csv>})
\\
\textit{name}\bigl(\texttt{<http://example.com/ro/123/survey.csv>},
\text{``Survey of care providers''}\bigr)
\\
\\
\textit{Dataset}(\texttt{<http://example.com/ro/123/interviews/>})
\\
\textit{name}\bigl(\texttt{<http://example.com/ro/123/interviews/>},
\\
\quad \text{``Audio recordings of care provider interviews''}
\bigr)
\\
\textit{license}(\texttt{<http://example.com/ro/123/interviews/>},
\\*
\quad \texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>})
\end{gather*}}}

>>>>

The possibilities of consuming\footnote{Some consideration is needed in processing of RO-Crates as
knowledge graphs, e.g. establishing absolute IRIs for files inside a
ZIP archive, detailed in the RO-Crate specification: \url
{https://www.researchobject.org/ro-crate/1.1/appendix/relative-uris.html}.} RO-Crate metadata with such
powerful tools gives another strong reason for using Linked Data as a
foundation. This use of mature Web\footnote{Note that an RO-Crate is not required to be published on the
Web, see Section~\ref{sec:selfdescribed}.} technologies also means its
developers and consumers are not restricted to the Research Object
aspects that have already been specified by the RO-Crate community, but
can extend and integrate RO-Crate in multiple standardised ways.

@@

One aim of RO-Crate is to be conceptually simple. This simplicity has
been repeatedly checked and confirmed through an informal community
review process. For instance, in the discussion on supporting ad-hoc
vocabularies\footnote{\url{https://github.com/ResearchObject/ro-crate/issues/71}} in
RO-Crate, the community explored potential Linked Data solutions. The
conventional wisdom in RDF best
practices\footnote{\url{https://www.w3.org/TR/swbp-vocab-pub/}} is to establish a
vocabulary with a new IRI namespace, formalised using RDF
Schema\footnote{\url{http://www.w3.org/TR/2014/REC-rdf-schema-20140225/}} or
OWL\footnote{\url{http://www.w3.org/TR/2012/REC-owl2-overview-20121211/}}
ontologies. However, this may seem an excessive learning curve for
non-experts in semantic knowledge representation, and the RO-Crate
community instead agreed on a dual lightweight approach: (i)
Document\footnote{\url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html\#adding-new-or-ad-hoc-vocabulary-terms}}
how projects with their own Web-presence can make a pure HTML-based
vocabulary, and (ii) provide a community-wide PID namespace under
\texttt{https://w3id.org/ro/terms} that redirect to simple CSV files
maintained in GitHub.\footnote{\url{https://github.com/ResearchObject/ro-terms}}

@@

RO-Crate reuses vocabulary terms and IRIs from Schema.org, but provides
its own versioned JSON-LD
context,\footnote{\url{https://w3id.org/ro/crate/1.1/context}} which has a flat list
with the mapping from JSON-LD keys to their IRI equivalents (e.g. key
\texttt{"author"} maps to the \url{http://schema.org/author} property).

@@

An $\textit{ROCrate}(R)$ is defined as a self-described \emph{Root Data Entity},
which contains parts (\emph{data entities}), which are
further described in \emph{contextual entities}. These terms align with
their use in the RO-Crate 1.1 terminology.\footnote{\url{https://www.researchobject.org/ro-crate/1.1/terminology}}

The $\textit{Root}(r)$ is a type of $\textit{Dataset}(r)$, and must as metadata 
have at least the attributes $\textit{name}$, $description$ and
$\textit{datePublished}$ as literal values, as well as a contextual entity that identify its
$\textit{license}$. These predicates correspond to the RO-Crate 1.1 minimal requirements
for the root data entity.\footnote{\url
{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html\#direct-properties-of-the-root-data-entity}}

@@

The below is an example RO-Crate represented using the above
formalisation, assuming a base IRI of \texttt{<http://example.com/ro/123/>}:
%
{\fontsize{9.7}{11.7}\selectfont{\begin{gather*}
\textit{ROCrate}(\texttt{<http://example.com/ro/123/>})
\\
\textit{name}\bigl(\texttt{<http://example.com/ro/123/>},
\\
\quad \text{``Data files associated with the manuscript:Effects of {
\ldots}''}\bigr)
\\
\textit{description}\bigl(\texttt{<http://example.com/ro/123/>},
\\
\quad \text{``Palliative care planning for nursing home residents {
\ldots}''}\bigr)
\\
\textit{license}(\texttt{<http://example.com/ro/123/>},
\\
\quad \texttt{<https://spdx.org/licenses/CC-BY-4.0>})
\\
\textit{datePublished}\bigl(\texttt{<http://example.com/ro/123/>}, \text
{``2017-02-23''}\bigr)
\\
\textit{hasPart}(\texttt{<http://example.com/ro/123/>}, \texttt{<http://example.com/ro/123/file.txt>})
\\
\textit{hasPart}(\texttt{<http://example.com/ro/123/>}, \texttt{<http://example.com/ro/123/interviews/>})
\\
\\
\textit{ContextualEntity}(\texttt{<https://spdx.org/licenses/CC-BY-4.0>})
\\
\textit{name}\bigl(\texttt{<https://spdx.org/licenses/CC-BY-4.0>},
\\
\quad \text{Creative Commons Attribution 4.0''}\bigr)
\\
\\
\textit{ContextualEntity}(\texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>})
\\
\textit{name}\bigl(\texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>},
\\
\quad \text{Creative Commons Attribution Non Commercial 4.0''}
\bigr)
\\
\\
\textit{File}(\texttt{<http://example.com/ro/123/survey.csv>})
\\
\textit{name}\bigl(\texttt{<http://example.com/ro/123/survey.csv>},
\text{``Survey of care providers''}\bigr)
\\
\\
\textit{Dataset}(\texttt{<http://example.com/ro/123/interviews/>})
\\
\textit{name}\bigl(\texttt{<http://example.com/ro/123/interviews/>},
\\
\quad \text{``Audio recordings of care provider interviews''}
\bigr)
\\
\textit{license}(\texttt{<http://example.com/ro/123/interviews/>},
\\*
\quad \texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>})
\end{gather*}}}
